Physics-based protein structure prediction and design using the confinement method

### 2013-01-08 Paper/paper.tex
merged comments from Arijit

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecular
conformations is currently a particularly difficult and computationally expensive task with existing
methods. In this work, an improved \Alberto{Arijit: what is the improvement??} version of the confinement approach is used to calculate absolute
free energies of bio molecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. Throught out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preference set the stage for the use of the method in
protein design.

@@

In computational structural biology there are numerous cases where free energy between well defined
states are necessary. Examples ranges from small to large conformational change of protein due to
ligand binding, change of pH etc \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also play an important role in case of protein
folding. The ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting point or unfolding
it by changing different condition most proteins will finally assume the same structure. It is very
important for proteins to achieve their native conformation since defects in protein folding may be
the molecular cause of a range of human genetic disorder. For example a misfolded protein known as
prion when enters any healthy organism it converts properly folded protein into misfolded one.
Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein. These observations further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change become difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescale involved in such tranformation may
be beyond timescale of direct classical molecular dynamics simulation.  Generally, in order to
calculate free energy a pathway or reaction coordinate is created with the intermediate structures.
Method like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and require large computational effort. The idea of reaction coordinate became more
complicated in case of protein folding.  As free energy landscape of the protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}
it can be trapped in a energy well during its visit in the conformational space. Even with the
advanced method like replica exchange molecular dynamics the three dimensional structure may be
trapped in a local minima for a considerable time. Due to various reasons it can also be misfolded
and trapped in a local energy minima. And one can mistakenly identify the misfolded state as the
native state (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

Calculation of free energy has been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}, . In recent years methods like Refrence system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random 
Work, Confinement Method have came out for calculation of free energy method.

Some of the groups
relies on one of the great strength of free energy calculation, that it is a state function and thus
it is independent of path.

In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. The free energy is a state function and thus it can be independent of path. 
Confinement method relies on this fact and uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraint to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy of conformation of pair of proteins with similar sequence but completely
different fold was calculated. Encouraged by the result, we applied the method of different targets
of CASP9 competition.

@@

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, We first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the prefential structure of the chamelon sequence. Another of the 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9 10, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimental structure (once it was available from the PDB). As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

\begin{table}
\label{table:casp_control}
\begin{center}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $12.49 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-2.48 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $22.00 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.24 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $22.15 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $-9.40 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}

@@

@@

@@

\subsection{The confinement method correctly identifies the most native-like predictions from a
subset of CASP predictions}
\Alberto{Arijit: we need a global table of success, rather than individual tables. The simplest form
    would have 3 columns: name of the PDB, number of structures tested, percent of success of
    picking 1st model}

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test world wide competition in which different groups apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. Our second test was to participate as a
metapredictor group in CASP10. In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

@@

\subsubsection{CASP9}

The question that we want to answer here is whether we can distinguish between native structure and the submitted models? 
%Identifying the native state from a set of models is a very easy test, that most metapredictors
%already handle correctly, but it is also a first proof of principle. 
First we choose a protein BVU3908 from Bacteroides vulgatus whose PDB id and CASP target code are
2L01 and T0559 respectively. The native structure of this 69 residue protein was solved using NMR.
The best predictor group for this target was "BAKER-ROSETTASERVER".
We initlally checked similarity between the models submitted by "BAKER-ROSETTASERVER" and discarded two of them from the 
analysis on the basis of being too similar to some of the other models. The GDT\_TS score and rmsd
values as shown in Figure~\ref{fig:T0559} indicate that the model 1 was predicted correctly, 
whereas the order of model 3 and 5 was wrong. The main difference
between the model 3 and the rest of the model is that the orientation of first alpha helix of model 3 is opposite.
On the other hand, as shown in Figure~\ref{fig:T0559} confine and release method not only can differentiate the native 
structure from the submitted models , it can also correctly rank all the models.

To further test the method we have taken the example of protein BT2368 from Bacteroides thetaiotaomicron. The PDB id of this 
74 amino acid residue protein is 2L02 and CASP target code is T0560. We have compared the free energy difference between 
the native structure and the two of the five submitted models from the group "Splicer". The remaining three models are discarded as 
again they are similar to the rest of the models. The comparison of GDT\_TS score as shown in 
Figure~\ref{fig:T0560} and published in final CASP result is only for models with residues 3 to 66. In order to keep
consistency, we have also done our analysis with models and crystal structure consisting residues 3 to 66. As expected,
the native state is identified correctly, and the two other models are ranked in the correct order.

@@

There are however some instances where the method can fail, specially when GDT scores are
very close. We have found this for an engineered protein from Asr4154 protein (PDB ID: 2L09 and CASP Target T0538).
This protein contains 54 amino acid residues with three alpha helixes and two very short beta strand. 
The model that is closest to the native structure was generated by the group PconsR with a GDT\_TS value of 96.23.
We have also considered one model from group "Shell" (GDT\_TS = 90.09) and "FOLDIT" (GDT\_TS = 86.32) for our analysis along with
the native crystal struture. The models from PconsR, Shell and FOLDIT are labelled as model 1, model 2 and model 3 
respectively in Figure~\ref{fig:T0538}. As seen in Figure~\ref{fig:T0538}, one can find that the 
model from PconsR has lowest free energy value, making it a more stable structure. Although ranking of the other models
remain consistent, it may happen that the theoretical model produced by the group "PconsR" much more stable than the 
crystal structure. This is particularly difficult target for rank ordering as all the models those are considered for this 
calculation are within RMSD values of 2.6 \AA.

There is another instance where this method fails. This is the case of solution NMR structure of a domain of adhesion exoprotein 
from Pediococcus pentosaceus. The
pdb id and CASP code of this 79 residue protein are 2KWY and T0569 respectively. We have compared the native structure with the
best model structure submitted for this target in CASP9. This model structute was submitted by the group " Mufold". ALl the
five model structures submitted by this group was found to be close. Thus we proceed by comparing only the native structure and
the best predicted model of this target. Visualization upon superimposition of two structures reveals that
the difference between the two structure is at the region consist of residues 48 to 65, where model1 is much more disordered compare to the
native structure(see Figure~\ref{fig:T0569}) . Other regions look similar for both the cases with of model1 is within 2.6 \AA
rmsd value of the native structure. In this case we found that the confinement method predict that the predicted model is stable than the
native structure with a free energy difference of $9.4 \pm 0.7$ Kcal/Mol.

@@

\subsubsection{Difficulty in CASP Experiments}

>>>>

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecular
conformations is particularly difficult and computationally expensive task with existing
methods. In this work, confinement approach is used to calculate absolute
free energies of bio molecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. The free energy calculated can be decomposed into per residue contribution in a 
approximate way. Per residue free energy allow us
to identify the reason behind conformational preference of the biomolecules. Through out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preference and per residue free energy decomposition set the stage for the use of this 
method in protein design.

@@

In computational structural biology there are numerous cases where free energy between well defined
states are necessary. Examples ranges from small to large conformational change of protein due to
ligand binding, change of pH etc \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also play an important role in case of protein
folding. The ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting point or unfolding
it by changing different condition most proteins will finally assume the same structure. It is very
important for proteins to achieve their native conformation since defects in protein folding may be
the molecular cause of a range of human genetic disorder. For example a misfolded protein known as
prion when enters any healthy organism it converts properly folded protein into misfolded one.
Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein. These facts further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change become difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescale involved in such tranformation may
be beyond timescale of direct classical molecular dynamics simulation.  Generally, in order to
calculate free energy a pathway or reaction coordinate is created with the intermediate structures.
Method like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and require large computational effort. The idea of reaction coordinate became more
complicated in case of protein folding.  As free energy landscape of the protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}
it can be trapped in a energy well during its visit in the conformational space. Even with the
advanced method like replica exchange molecular dynamics the three dimensional structure may be
trapped in a local minima for a considerable time and trapped in a local energy minima
and one can mistakenly identify the misfolded state as the native state.
% (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

Calculation of free energy has been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}, . In recent years methods like Refrence system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008}, Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} have came out for calculation of free energy method. 
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that free energy is a state function and 
thus independent of path. It uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraint to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy of conformation of pair of proteins with similar sequence but completely
different fold was calculated. Encouraged by the result, we applied the method of different targets
of CASP9 competition and tested whether it can predict the most stable conformation of the protein based 
on its free energy. The remarkable ability of confinement method to identify residues which 
stabilizes/destabilizes a particular conformation from per residue free energy analysis set the stage
for future application of this method in protein design.

@@

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, We first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the prefential structure of the chamelon sequence. Other 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimentally determined structure. As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
Most interestingly, we decomposed the free energy into its per residue component, which help 
us to identify the residues which stabilize or destabilize a particular conformation of the protein.
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}

@@

\subsection{Per residue free energy calculation can identify the mechanistic detail behind conformational pereference 
of a particular residue}

We decomposed the calculated free energy into per residue decomposition as described in method section. This help us for
identification of important residues that stabilizes a particular conformation. 
In this direction, We first calculate the per residue free energy for the GA95 and GB95 sequences. Similar to the total 
free energy for each sequence, we calculate the free energy difference between $4 \beta + \alpha$ and $3 \alpha$ structure for each 
residue. The free energy, $\Delta \Delta G (4 \beta + \alpha$ - $3 \alpha)$ is plotted in Figure~\ref{fig:orban_full}(F). 
In this plot, the negative peak indicate the residue will stabilize the $4 \beta + \alpha$ conformation of a particular
residue and the positive peak will stabilize the $3 \alpha$ structure. As, the differences of amino acid 
residues are only at 3 position, there are some common features of residues in both GA95 and GB95 sequences in both 
$3 \alpha$ and $4 \beta + \alpha$ conformation. They are reported in Figure~\ref{fig:orban_full} 
and Table~\ref{table:orban_perresidue}. Before, we further go into that detail, it is important to note that 
experimental observations classified the protein into two parts: Amino acids 9-51 are fully structured in both
the fold, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favors $4 \beta + \alpha$ structure and can also seen from the plot 
Figure~\ref{fig:orban_full}(F). Here, we try to explain some of the largest peak in that plot. The residues Leu-5 and 
Leu-7 favors $4 \beta + \alpha$ structure as these hydrophobic residues are oriented towards the protein hydrophobic 
core, whereas in $3 \alpha$ structure they are exposed to the solvent (Figure~\ref{fig:orban_full}(A)). One can argue that 
hydrophilic residues in these positions can stabilize the $3 \alpha$ conformation.
The next big peak is located at position 11. A close observation of the sidechain orientation of residue Gln at position 11 reveal that it
is exposed to the protein surface in both the conformation, but in $3 \alpha$ structure it has an additional H-bond with Glu-15 which gives
it more stability compare to the $4 \beta + \alpha$ structure (Figure~\ref{fig:orban_full}(B)). The peak 
at position 14 is due to the hydrophilic Glu, which is completely
exposed to the solvent in $4 \beta + \alpha$ conformation as shown in (Figure~\ref{fig:orban_full}(E)), but only partially exposed 
to the solvent in $3 \alpha$ form. 
This is the reason why Glu at position 14 stabilize $4 \beta + \alpha$ conformation. Another high peak is at position 26, which indicate
that the residue in that position stabilizes $4 \beta + \alpha$ conformation. Again, the hydrophobic amino acid residue, Ala at position 26
is oriented towards the hydrophobic core of the protein in $4 \beta + \alpha$ conformation , whereas it is exposed towards the protein surface 
in $3 \alpha$ structure (Figure~\ref{fig:orban_full}(C)). Similar situations arise for residues at position 54 and 56. 
These residues are in the region, where $4 \beta + \alpha$ form is fully structured and $3 \alpha$ form is unstructured. 
As shown in Figure~\ref{fig:orban_full}(D), the sidechain of hydrophobic Val-54 is oriented towards the hydrophobic core in 
$4 \beta + \alpha$ conformation and thus have hydrophobic interactions with other residues in this conformation.
On the other hand, in $3 \alpha$ it is exposed to the solvent. The reverse happens for the residue at position 56.
Here the hydrophilic residue Glu is partially or not at all exposed to the protein surface in $4 \beta + \alpha$ but fully exposed 
to the solvent in $3 \alpha$ conformation (Figure~\ref{fig:orban_full}(G)). This is the reason, why one can see a positive peak for position 56. 
Similarly, hydrophobic Ile-49 in $4 \beta + \alpha$ conformation is exposed to protein surface and solvent, whereas as shown in 
Figure~\ref{fig:orban_full}(H)$3 \alpha$ conformation 
it has hydrophobic interaction with other hydrophobic residues in protein core.   
Now, let us discuss the differences of $3 \alpha$ and $4 \beta + \alpha$ fold due to specific difference of mutated amino acid 
residues in GA and GB sequence. The differences are in positions 20, 30 and 45. While GA95 has Leu-20-Ile-30-Leu-45, GB95 sequence 
has Ala-20-Phe-30-Tyr-45 at those position. One can find that the peak at position 49 stabilizes the GA sequence more than that of GB
sequence. In $3 \alpha$ fold, Ile-49 has hydrophobic interactions with Leu-20 and Ile-30 with GA sequence, whereas with GB sequence it 
has such interactions with smaller Ala-20 and Phe-30 residues (Figure~\ref{fig:orban_full}(I). In $3 \alpha$, with GB sequence, Phe-30 
being a larger group can not be completely accomodated in the hydrophobic core(Figure~\ref{fig:orban_full}(J)) and therefore, 
partially exposed to the protein surface. There is a very interesting difference for residue at position 45 as well. In GB95 sequence
and $4 \beta + \alpha$ fold there is H-bonding between Tyr-45 and Asp-47 (Figure~\ref{fig:orban_full}(K)), which is absent in the
$3 \alpha$ conformation. On the other hand with GA conformation, this H-bond is absent for both the conformation as there is a Leu at position
45. Although, one can not find much difference between $4 \beta + \alpha$ and $3 \alpha$ conformation
for position 20 for both GA95 and GB95 sequence. There are significant difference for the residues at position 30 and 45.  
These residues have some long term effects as well. As already explained, residue Ile-49 has much more hydrophic interactions with residues
at position 20 and 30 in GA sequence, which is less in GB sequence. There are difference for Asp-47 as well. With $3 \alpha$ conformation 
and GA95 sequence, Asp-47 has H-bonding with Lys-50, which is absent in the other conformation.

\begin{table}
\begin{center}
\caption{Analysis of per residue free energy decomposition of GA95 and GB95 sequence reveal that certain residues prefer
either $3 \alpha$  or $4 \beta + \alpha$ structure. These observations and the reason behind such conformational preferences are listed in this
table.}
\label{table:orban_perresidue}
\begin{tabular}{| l | l | l |}
\hline
    Observation                                & $3 \alpha$ structure           &    $4 \beta + \alpha$ structure                       \\ \hline
    Leu-5 and Leu-7 favor                      & hydrophobic residues are       & oriented towards the                                  \\ 
    $4 \beta + \alpha$                         & exposed to the solvent         & hydrophobic core                                      \\ \hline
    Gln-11 favors $3 \alpha$                   & Gln-11 has H-bond with Glu-15  & no such H-bond                                        \\ \hline
    Glu-14 favors $4 \beta + \alpha$           & hydrophilic residue partially  & completely exposed                                    \\       
                                               & exposed to the solvent         & to the solvent                                        \\ \hline        
    Ala-26 favors  $4 \beta + \alpha$          & hydrophobic residue oriented   & oriented towards the                                   \\ 
                                               & towards surface                & hydrophobic core                                       \\ \hline
    Ile-49 favors $3 \alpha$                   & hydrophobic interaction        & exposed to the                                          \\
                                               & inside the protein             & solvent                                                \\ \hline 
    Val-54 favors $4 \beta + \alpha$           & hydrophobic residue exposed    & hydrophobic interaction                                \\
                                               & to the solvent                 & inside the protein                                      \\ \hline
    Glu-56 favors $3 \alpha$                   & hydrophilic residue partially  & completely exposed                                    \\
                                               & exposed to the solvent         & to the solvent                                        \\ \hline
    Phe-30 in GB sequence                      & exposed to the solvent         & oriented towards the                                  \\
    favors $4 \beta + \alpha$                  & in protein surface             & hydrophobic core                                      \\ \hline   
    Tyr-45 in GB Sequence                      & No H-bonding                   & H-bond with Asp-47                                    \\  
    favors $4 \beta + \alpha$                  &                                &                                                       \\ \hline
    Asp-47 in GA sequence                      & H-bond with Lys-50             & No H-bond, No Tyr-45                                  \\  
    favors $3 \alpha$                                 &                         & in GA sequence                                         \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}
\includegraphics[width=7 in,height=5.8 in]{orban_full.pdf}
\label{fig:orban_full}
\caption{Orientation of residues of different sidechains in $4 \beta + \alpha$ and $3 \alpha$ form
A. Orientation of Hydrophobic residues Leu-5 and Leu-7.
B. Gln-11 has H-bond with Glu-15 in $3 \alpha$, which is absent in $4 \beta + \alpha$
C. Hydrophobic Ala-26
D. Hydrophobic Val-54.
E. Hydrophilic Glu-14.
F. The plot of free energy difference for GA95 and GB95 sequence. Positive peak stabilize $3 \alpha$ form and negative peak 
stabilize $4 \beta + \alpha$. 
G. Hydrophilic Glu-56 .
H. Orientation of Hydrophic Ile-49.
I. With GA95 sequence, Ile-49 have hydrophic interaction with both Leu-20 and Ile-30.
J. Larger Phe-30 in GB sequence, move towards the protein surface in $4 \beta + \alpha$.
K. Tyr-45 form H-bond with Asp-47 in $4 \beta + \alpha$.}  
\end{figure}

@@

\subsection{The confinement method correctly identifies the most native-like predictions from a
subset of CASP predictions}

%\Alberto{Arijit: we need a global table of success, rather than individual tables. The simplest form
%    would have 3 columns: name of the PDB, number of structures tested, percent of success of
%    picking 1st model}

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test in which different groups world wide apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. We have tested different options for that purpose.
First, we rank order submition from same group. Then we rank submition from different groups. Our second test was 
to check whether this method can do better quality assessment/metaprediction than the best group in that catagory in CASP9. 
In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

@@

\subsubsection{confinement method can rank different models submitted by the same group and distinguish the native structure}

The question that we want to answer here is whether we can rank models submitted by the same group.In all cases, we have calculated the free energy
difference between the native structure and the submitted models as well. 
%Identifying the native state from a set of models is a very easy test, that most metapredictors
%already handle correctly, but it is also a first proof of principle. 
First we choose a protein BVU3908 from Bacteroides vulgatus whose PDB id and CASP target code are
2L01 and T0559 respectively. The native structure of this 69 residue protein was solved using NMR.
The best predictor group for this target was "BAKER-ROSETTASERVER".
We initlally checked similarity between the models submitted by "BAKER-ROSETTASERVER" and discarded two of them from the 
analysis on the basis of being too similar to some of the other models. The GDT\_TS score and rmsd
values as shown in Figure~\ref{fig:T0559} indicate that the model 1 was predicted correctly, 
whereas the order of model 3 and 5 was wrong. The main difference
between the model 3 and the rest of the model is that the orientation of first alpha helix of model 3 is opposite.
On the other hand, as shown in Figure~\ref{fig:T0559} confine and release method not only can differentiate the native 
structure from the submitted models , it can also correctly rank all the models like their The GDT\_TS score.

To further test the method we have taken the example of protein BT2368 from Bacteroides thetaiotaomicron. The PDB id of this 
74 amino acid residue protein is 2L02 and CASP target code is T0560. We have compared the free energy difference between 
the native structure and the two of the five submitted models from the group "Splicer". The remaining three models are discarded as 
again they are similar to the rest of the models. The comparison of GDT\_TS score as shown in 
Figure~\ref{fig:T0560} and published in final CASP result is only for models with residues 3 to 66. In order to keep
consistency, we have also done our analysis with models and crystal structure consisting residues 3 to 66. As expected,
the native state is identified correctly, and the two other models are ranked in the correct order in accordance with the GDT\_TS score.

\subsubsection{Confinement method can rank models submitted by different groups}

@@

\subsubsection{Per residue free energy calculation identify the residues that are responsible for difference between two conformation}

In order to understand the per residue contribution towards the free energy, we have first calculated the total free energy of NMR 
structure of a domain of adhesion exoprotein from Pediococcus pentosaceus and its submitted CASP9 model. The 
pdb id and CASP code of this 79 residue protein are 2KWY and T0569 respectively. We have compared the native structure with the
best model structure submitted for this target in CASP9. This model structure was submitted by the group " Mufold". All the
five model structures submitted by this group was found to be close, within a GDT\_TS score  of 72 to 78. Thus we proceed by 
comparing only the native structure and
the best predicted model of this target. Visualization upon superimposition of two structures reveals that
the difference between the two structure is at the region consist of residues 48 to 65, where model1 is much more disordered compare to the
native structure(see Figure~\ref{fig:T0569}). Other regions look similar for both the cases with of model1 is within 2.6 \AA backbone
rmsd value of the native structure. We found that, confinement method can still differentiate between the native structure and 
the generated model. We have also calculated the per residue free energy with a aim that this can identify the residue which are responsible
for such free enerfy difference. The per residue free energy decomposition is shown in  Figure~\ref{fig:T0569_per_residue}(A). We indeed 
found that two hydrophobic residues Val-59 and Ile-61 are oriented towards the solvent in the protein 
surface in the generated model (Figure~\ref{fig:T0569_per_residue}(B). This the region, where there is a beta sheet in the original crystal structure, which is absent in the 
model. There are also some region, which stabilizes the generated model compare to that of the crystal. A close inspection reveal that 
the hydrophilic Glu in position 67 is completely exposed to the protein surface and solvent in the best model, whereas it is partially exposed in the 
crystal. This is presented in Figure~\ref{fig:T0569_per_residue}(C). The peak at position 76 is due to the H-bond between Lys-76 and Asp-11, 
which is absent in original crystalographic structure (Figure~\ref{fig:T0569_per_residue}(D)).

\subsubsection{Confinement method fails ?}

There are however some instances where the method can fail, specially when GDT scores and all the structures are
very close. We have found this for an engineered protein from Asr4154 protein (PDB ID: 2L09 and CASP Target T0538).
This protein contains 54 amino acid residues with three alpha helixes and two very short beta strand. 
The model that is closest to the native structure was generated by the group PconsR with a GDT\_TS value of 96.23.
We have also considered one model from group "Shell" (GDT\_TS = 90.09) and "FOLDIT" (GDT\_TS = 86.32) for our analysis along with
the native crystal struture. The models from PconsR, Shell and FOLDIT are labelled as model 1, model 2 and model 3 
respectively in Figure~\ref{fig:T0538}. As seen in Figure~\ref{fig:T0538}, one can find that the 
model from PconsR has lowest free energy value, making it a more stable structure. Although ranking of the other models
remain consistent, it may happen that the theoretical model produced by the group "PconsR" much more stable than the 
crystal structure. This is particularly difficult target for rank ordering as all the models those are considered for this 
calculation are within RMSD values of 1.6 - 2.0 \AA. In order to understand the reason behind the mismatch of GDT\_TS values
of crystal and the model1, we have decomposed the total free energy into the per residue component. A close analysis of 
the per residue free energy value, we found that although, crystal and different model have closed GDT\_TS and backbone RMSD 
value, there are number of sidechains oriented in a very different way. As shown in Figure~\ref{fig:T0538compare}(A),
there is a salt bridge between Arg-32 and Glu-35 present in the crystallographic structure, which is absent in the
model 1. In model 1, Arg-32 has another salt bridge with Glu-28. Glu-28 in crystal has the salt bridge with Lys-24. In model 1, Lys-24 is
exposed towards the protein surface. This is shown in Figure~\ref{fig:T0538compare}(A). There are more such interesting and may be crucial
sidechain differences between the model and the crystallographic structure, which indicate the limitation of the GDT\_TS based 
ranking of the protein structure. As, shown in Figure~\ref{fig:T0538compare}(B), Arg-26 and Glu-50 has a salt bridge in the crystallographic 
structure. Also, Phe-51 is oriented towards the protein core in the crystallographic structure. In model 1, the salt bridge between Arg-26 and Glu-50
is absent and the Phe-51 is oriented towards the protein surface.

@@

%\subsubsection{Difficulty in CASP Experiments}

%It is often difficult to model CASP targets using the confinement and release method. There are many cases,
%where the sequence of a particular target is trimmed during actual result presentation. This is illustrated
%with the crystal structure of the leucine zipper domain of cGMP dependent protein kinase I (pdb id: 3NMD
%and CASP9 code: T0605) as shown in Figure~\ref{fig:T0605}. The actual CASP 9 target consisted of 72 amino acid residues, but
%as only 18-66 amino acid residues could be detected in actual crystal structure of this protein, the final result was trimmed 
%with respect to that crystal structure with 49 amino acid residues which consist of a single alpha helix. The amino acid residues 18 and 66 are shown in
%three models submitted by the group "Baker". It is important to note that blind prediction of model ranking on the basis of free energy can be hugely 
%affected in such scenario as the only major difference between those models are the trimmed parts, specially the region consisting residues 1-18.
